Restrictive Clearance Comparison

### 2024-12-19 data-raw/RestrictiveClearance_comparison.Rmd
Added uncertainty estimation section to data-raw/RestrictiveClearance_comparison.Rmd and updated a few of the visualizations.

```{r library_loading}
devtools::load_all()
library(tidyverse)
my_pk <- readRDS("data-raw/all_cvt_pk.rds")
source("data-raw/RC_comparison_bkgd.R") # httk loaded here
```

@@

1. Get chemicals + species combinations in CvTdb that have _httk_ predictions.
2. Safely get parameters, clearance, and solutions for each of these groups.
3. Compare clearances from _httk_ with those from _invivoPKfit_.
4. Compare $f_{up}$ for subset of data with total values available via _httk_.
5. Use new _clearance-aware_ optimization of 1-compartment parameters to
benchmark _httk_ solutions.
6. Estimate the fraction unbound in plasma for chemicals with rat data,
for those same chemicals with human data for Clint and Fup. 
Begin with the assumption that there is restrictive clearance, those chemicals
that do not change when optimized (and produce reasonable fits) are likely to
be _restrictively cleared_, whereas those that increase Fup during maximum
likelihood estimation are _possibly non-restrictively clearaed_. 
7. Other factors such as renal clearance may play a role in those cases.
Validate the clearance estimate by measuring changes in predicted clearance.
Important to identify whether there seems to be a "fraction missing".
8. Wrap the previous steps into attempts to identify true positive
categorizations for Restrictive and Potentially Non-Restrictively metabolized
chemicals (include Lipinski's Rule of 5 too).

@@

```{r clearance_viz}
combined_clearance_df %>%
  mutate(AD_restrictive = abs(CLtot.tkstats - restrictive_clearance),
         AD_nonrestrictive = abs(CLtot.tkstats - nonrestrictive_clearance),
         rough_determination = ifelse(
           min(AD_restrictive, AD_nonrestrictive) == AD_restrictive &
             CLtot.tkstats < nonrestrictive_clearance,
           "Restrictive", "Possibly non-Restrictive"
         )) %>% 
  dplyr::distinct(Chemical, rough_determination) %>%
  tally()

@@

ccd %>%
  ggplot(aes(
    x = Value,
    y = Chemical,
    fill = Clearance_Source
  )) +
  geom_point(shape = 21, alpha = 0.7) +
  scale_x_log10() +
  scale_fill_manual(
    breaks = c(
      "nonrestrictive_clearance", "CLtot.tkstats",
      "CLtot.nca_mean",
      "restrictive_clearance"
    ),
    labels = c(
      "non-restrictive clearance (httk)",
      "predicted clearance (invivoPKfit)",
      "mean non-compartmental total clearance (invivoPKfit)",
      "restrictive clearance (httk)"
    ),
    values = c("orange2", "grey5", "skyblue3", "green4")
  ) +
  facet_grid(rows = vars(Species), scales = "free", space = "free") +
  labs(
    x = "Total Clearance",
    y = "Chemical"
  ) +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
    panel.background = element_blank(),
    panel.grid.major = element_line(color = "grey90", linewidth = 0.5),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(face = "bold", angle = 0),
    # axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "right",
    legend.direction = "vertical",
    axis.ticks = element_blank(),
    axis.line = element_blank()
  )

@@

ggsave(
  filename = paste0(
    Sys.getenv("FIG_DIR"),
    "rat_tkClearance_comparison_httk_PBPK.png"
  ),
  height = 8,
  width = 8,
  device = "png"
)
```

@@

fup_df %>%
  ggplot(aes(x = fup)) +
  geom_histogram(
    # aes(y = after_stat(density)),
    binwidth = 0.03,
    color = "grey5", fill = "grey5"
  ) +
  geom_histogram(
    data = fup_df %>%
      filter(Chemical %in% common_chems_rat),
    # aes(y = after_stat(density)),
    fill = NA,
    color = "darkgreen", binwidth = 0.03
  ) +
  expand_limits(expand = c(0,0)) +
  # geom_vline(xintercept = 0.05, linetype = "dashed") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  scale_x_continuous(expand = expansion(mult = c(0.01, 0.01))) +
  labs(
    title = "Distribution of Funbound.plasma values",
    subtitle = "(chemicals fit by invivoPKfit in dark green)"
  ) +
  coord_cartesian(clip = "off") +
  theme_bw()

@@

# Functions wrapping httk single dose evaluation
httk_bolus_gas_pbtk <- function(dtxsid, species = "Human",
                                dose, times, iv.dose, LOQ) {
  tmp_solution <- solve_gas_pbtk(
    dtxsid = unique(dtxsid),
    dose = unique(dose),
    exp.conc = 0,
    iv.dose = unique(iv.dose),
    times = unique(c(0, times/24)),
    input.units = "mg/kg",
    output.units = "mg/L",
    species = unique(stringr::str_to_title(species)),
    default.to.human = TRUE,
    suppress.messages = TRUE
  )
  # Remove the 0.0001 intermediate time value that is returned
  retdf <- tibble(
    Time = round(tmp_solution[, "time"]*24, digits = 3),
    httk_Preds = tmp_solution[, "Cplasma"]
  )
  retdf[which(retdf$Time %in% unique(round(times, 3))), ]
}
```

@@

```{r rat_data}
my_data <- get_data(my_pk)

rat_df <- my_data %>%
  distinct(Chemical, Species, Dose, Route, Media, Time, Conc, exclude, Detect,
           pLOQ) %>%
  filter(Species %in% "rat", Media %in% "plasma",
         Chemical %in% common_chems_rat, !exclude, Detect) %>%
  mutate(iv_dose = (Route == "iv"),
         Time = round(Time, 3)) %>%
  group_by(Chemical, Species, Dose, Route, Media) %>%
  filter(n() > 4) %>%
  nest(.key = "CvT") %>%
  rowwise() %>%
  mutate(
    httk_solved = list(
      tryCatch(
        expr = do.call(httk_bolus_gas_pbtk,
                       list(
                         dtxsid = Chemical,
                         species = Species,
                         dose = Dose,
                         times = CvT$Time,
                         iv.dose = CvT$iv_dose,
                         LOQ = CvT$pLOQ
                       )
        ) %>%
          inner_join(tibble(Time = CvT$Time, Conc = CvT$Conc, pLOQ = CvT$pLOQ),
                     by = join_by(Time)),
        error = function(err) {
          message(paste0("Unable to get solution for ", Chemical))
          print(err)
        } 
      )
    )
  ) %>% filter(Chemical %in% clear_chem_levels)

@@

rat_df <- rat_df %>%
  dplyr::select(Chemical, Species, Dose, Route, Media, httk_solved) %>%
  unnest(cols = httk_solved)

rat_out <- rat_df %>%
  mutate(httk_Preds = ifelse(httk_Preds < pLOQ, pLOQ, httk_Preds)) %>% 
  mutate(SLE = (log2(Conc) - log2(httk_Preds))^2,
         AE = abs(Conc - httk_Preds)) %>%
  summarize(
    RMSLE = sqrt(mean(SLE, na.rm = TRUE)),
    MAE = mean(AE, na.rm = TRUE)
   ) %>% pivot_longer(cols = c(RMSLE, MAE),
                      names_to = "metric",
                      values_to = "value")

@@

rat_params %>%
  ggplot(aes(y = fupClint)) + 
  geom_point(aes(x = Clint), color = "darkred", shape = 1) +
  geom_abline(slope = 1, intercept = 0, color = "darkblue",
              linetype = "dashed")

rat_params %>%
  mutate(fup_z = scale(Funbound.plasma),
         clint_z = scale(Clint)) %>%
  dplyr::select(Chemical, fup_z, clint_z) %>%
  pivot_longer(cols = c(fup_z, clint_z),
               names_to = "pars", values_to = "vals") %>%
  ggplot(aes(x = pars, y = vals, group = Chemical)) +
  geom_line()

@@

```{r running-invivoPKfit}
# This is now updated to use dawson chem table
minimal_pk <- pk(data = cvt %>%
                   filter(analyte_dtxsid %in% common_chems_rat,
                          species %in% "rat"))

@@

new_pk <- do_fit(new_pk, n_cores = 14)
gc()

@@

@@

rel_ll %>%
  rowwise() %>% 
  mutate(across(contains("_1comp"),
                \(x) {
                  ifelse(
                    x < model_flat | x < 1,
                    NA_real_,
                    x
                  )
                })) %>% 
  drop_na(model_1comp_fup)

@@

new_pk$fit %>%
  filter(param_name %in% "Fup",
         model %in% "model_1comp_fup") %>%
  distinct(Chemical, Species, method, model,
           param_name, estimate, start) %>%
  mutate(direction = ifelse(abs(estimate - start) < 0.1,
                            "similar (less than 10% difference)",
                            ifelse(estimate > start,
                                   "increase",
                                   "decrease"))) %>%
  pivot_longer(cols = c(estimate, start),
               names_to = "parameterization",
               values_to = "value") %>%
  mutate(parameterization = factor(parameterization,
                                   levels = c("start", "estimate"))) %>%
  ggplot(aes(x = parameterization, y = value)) +
  geom_point() +
  geom_line(aes(group = Chemical)) +
  facet_grid(cols = vars(direction)) +
  labs(title = "Optimization of fraction unbound in plasma") +
  theme_bw()

ggsave(filename = paste0(
  Sys.getenv("FIG_DIR"),
  format(today(), '%d%b%Y'),
  "fup_optimization_generalFits"
))

@@

```{r new_analysis}
# Loading the new data
new_pk <- readRDS("data-raw/someCVT_wHTTK_JOINT.rds")

@@

## New reformulated models that estimate Clint!

@@

>>>>

```{r library_loading}
today <- format(Sys.Date(), "%d%b%Y")
devtools::load_all()
library(tidyverse)
library(ctxR)
my_pk <- readRDS("data-raw/all_cvt_pk.rds")
source("data-raw/RC_comparison_bkgd.R") # httk loaded here
```

@@

1. Get chemicals + species combinations in CvTdb that have _httk_ predictions.
2. Safely get parameters, clearance, and solutions for each of these groups.
3. Compare clearances from _httk_ with those from _invivoPKfit_.
4. Compare $f_{up}$ for subset of data with total values available via _httk_.
5. Use new _clearance-aware_ optimization of 1-compartment parameters to
benchmark _httk_ solutions.
6. Estimate the fraction unbound in plasma for chemicals with rat data,
for those same chemicals with human data for Clint and Fup. 
Begin with the assumption that there is restrictive clearance, those chemicals
that do not change when optimized (and produce reasonable fits) are likely to
be _restrictively cleared_, whereas those that increase Fup during maximum
likelihood estimation are _possibly non-restrictively cleared_. 
7. Other factors such as renal clearance may play a role in those cases.
Validate the clearance estimate by measuring changes in predicted clearance.
Important to identify whether there seems to be a "fraction missing".
8. Wrap the previous steps into attempts to identify true positive
categorizations for Restrictive and Potentially Non-Restrictively metabolized
chemicals (include Lipinski's Rule of 5 too).
9. Implement a _slow_ PBPK model that optimizes $f_{up}$ and $CL_{int}$ under
restrictive and non-restrictive clearance conditions. Test whether these have a
substantial difference from initial values. Use `coef_sd()` to evaluate
uncertainty of these parameter estimates for a larger share of chemicals.
10. Compare the results of 9 with the putative/predicted clearance from 1-6.

@@

```{r clearance_viz}
clearance_guess <- combined_clearance_df %>%
  mutate(AD_restrictive = abs(CLtot.tkstats - restrictive_clearance),
         AD_nonrestrictive = abs(CLtot.tkstats - nonrestrictive_clearance),
         rough_determination = ifelse(
           min(AD_restrictive, AD_nonrestrictive) == AD_restrictive &
             CLtot.tkstats < nonrestrictive_clearance,
           "Restrictive", "Possibly non-Restrictive"
         )) %>% 
  dplyr::ungroup()
clearance_guess %>%
  dplyr::distinct(Chemical, rough_determination) %>%
  group_by(rough_determination) %>%
  tally()

@@

ccd %>%
  ggplot(aes(
    x = Value,
    y = Chemical,
    fill = Clearance_Source
  )) +
  geom_point(shape = 21, alpha = 0.7) +
  scale_x_log10() +
  scale_fill_manual(
    breaks = c(
      "nonrestrictive_clearance", "CLtot.tkstats",
      "CLtot.nca_mean",
      "restrictive_clearance"
    ),
    labels = c(
      "non-restrictive clearance (httk)",
      "predicted clearance (invivoPKfit)",
      "mean non-compartmental total clearance (invivoPKfit)",
      "restrictive clearance (httk)"
    ),
    values = c("orange2", "grey5", "skyblue3", "green4")
  ) +
  facet_grid(rows = vars(Species), scales = "free", space = "free") +
  labs(
    x = "Total Clearance",
    y = "Chemical",
    title = "Comparing clearance estimates of chemicals with rat CvT data"
  ) +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
    panel.background = element_blank(),
    panel.grid.major = element_line(color = "grey90", linewidth = 0.5),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(face = "bold", angle = 0),
    # axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "right",
    legend.direction = "vertical",
    axis.ticks = element_blank(),
    axis.line = element_blank()
  )

@@

ggsave(
  filename = paste0(
    Sys.getenv("FIG_DIR"),"RestrictiveClearance_Analysis/",
    today,
    "rat_tkClearance_comparison_httk_PBPK.png"
  ),
  height = 8,
  width = 8,
  device = "png"
)
```

@@

fup_df %>%
  ggplot(aes(x = fup)) +
  geom_histogram(
    # aes(y = after_stat(density)),
    binwidth = 0.03,
    color = "grey5", fill = "grey5"
  ) +
  geom_histogram(
    data = fup_df %>%
      filter(Chemical %in% rat_pars_rat$param_df$Chemical),
    # aes(y = after_stat(density)),
    fill = NA,
    color = "darkgreen", binwidth = 0.03
  ) +
  expand_limits(expand = c(0,0)) +
  # geom_vline(xintercept = 0.05, linetype = "dashed") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  scale_x_continuous(expand = expansion(mult = c(0.01, 0.01))) +
  labs(
    title = "Distribution of Funbound.plasma values",
    subtitle = "(chemicals fit by invivoPKfit in dark green)"
  ) +
  coord_cartesian(clip = "off") +
  theme_bw()

@@

# Functions wrapping httk single dose evaluation
httk_bolus_gas_pbtk <- function(dtxsid, species = "Human",
                                dose, times, iv.dose, LOQ) {
  tmp_solution <- solve_gas_pbtk(
    dtxsid = unique(dtxsid),
    dose = unique(dose),
    exp.conc = 0,
    iv.dose = unique(iv.dose),
    times = unique(c(0, times/24)),
    input.units = "mg/kg",
    output.units = "mg/L",
    species = unique(stringr::str_to_title(species)),
    default.to.human = TRUE,
    suppress.messages = TRUE
  )
  # Remove the 0.0001 intermediate time value that is returned
  retdf <- tibble(
    Time = round(tmp_solution[, "time"]*24, digits = 3),
    httk_Preds = tmp_solution[, "Cplasma"]
  )
  retdf[which(retdf$Time %in% unique(round(times, 3))), ]
}

httk_preds_rat <- get_httk_preds(rat_pars_rat, my_pk, species = "rat")

@@

```{r rat_data}
rat_df <- httk_preds_rat %>%
  left_join(predict(my_pk)) %>%
  semi_join(get_winning_model(my_pk))

rat_df <- ggplot(mapping = aes(
    x = Conc_est,
    y = httk_Preds,
    color = interaction(Route, Media)
  )) +
  geom_point() +
  geom_line(mapping = aes(
    group = interaction(Chemical, Media, Route, Reference)
  )
  ) +
  geom_abline(slope = 1, intercept = 0, color = "black") +
  coord_equal(ylim = c(0, 200), xlim = c(0, 200))

rat_plots <- rat_df %>%
  distinct(Chemical, Species, Dose, Route, Media, Reference,
           Time, httk_Preds, Conc, Conc_est, pLOQ) %>%
  mutate(Dose = as.factor(Dose)) %>%
  group_by(Chemical, Species, Route, Media, Reference) %>%
  nest(.key = "data_pl")

rat_plots <- rat_plots %>%
  reframe(plots = purrr::map(data_pl,
                             \(x) {
                               cowplot::plot_grid(
                               ggplot(data = x %>%
                                        group_by(Time, Dose) %>%
                                        mutate(meanConc = mean(Conc,
                                                           na.rm = TRUE)) %>%
                                        select(-Conc) %>% distinct() %>%
                                        pivot_longer(cols = c(httk_Preds,
                                                              meanConc,
                                                              Conc_est),
                                                     names_to = "Measurement",
                                                     values_to = "Concentration"
                                                     ), 
                                      mapping = aes(
                                        x = Time,
                                        y = Concentration,
                                        color = Measurement,
                                        shape = Dose
                                      )) +
                                 geom_point() +
                                 geom_line(
                                   aes(group = interaction(Measurement, Dose))
                                   ) +
                                 theme_bw() +
                                 labs(
                                   title = paste(Chemical, Species),
                                   subtitle = paste(Route, Media)
                                 ),
                               ggplot(data = x) +
                                 geom_point(aes(x = Conc_est, httk_Preds,
                                                shape = Dose)) +
                                 geom_abline(slope = 1, intercept = 0,
                                             linetype = "dashed") +
                                 theme_bw()
                               )
                             })
  )

pdf(file = paste0(Sys.getenv("FIG_DIR"), "RC_comparison_CvTplots.pdf"),
    onefile = TRUE,
    height = 7,
    width = 10,
    bg = "white")

for (i in 1L:NROW(rat_plots)) {
  message("Now plotting: ", i)
  print(rat_plots %>% pull(plots) %>% .[[i]])
}

dev.off()

@@

rat_out <- rat_df %>%
  filter(Chemical %in% clear_chem_levels) %>%
  group_by(Chemical, Species, Route, Reference, Media) %>%
  mutate(httk_Preds = ifelse(httk_Preds < pLOQ, pLOQ, httk_Preds)) %>% 
  mutate(SLE = (log2(Conc) - log2(httk_Preds))^2,
         AE = abs(Conc - httk_Preds)) %>%
  summarize(
    RMSLE = sqrt(mean(SLE, na.rm = TRUE)),
    MAE = mean(AE, na.rm = TRUE)
   ) %>% pivot_longer(cols = c(RMSLE, MAE),
                      names_to = "metric",
                      values_to = "value")

@@

rat_params %>%
  ggplot(aes(y = fupClint)) + 
  geom_point(aes(x = Clint), color = "darkred", shape = 1) +
  geom_abline(slope = 1, intercept = 0, color = "darkblue",
              linetype = "dashed")
rat_params %>%
  mutate(fup_z = scale(Funbound.plasma),
         clint_z = scale(Clint)) %>%
  dplyr::select(Chemical, fup_z, clint_z) %>%
  pivot_longer(cols = c(fup_z, clint_z),
               names_to = "pars", values_to = "vals") %>%
  ggplot(aes(x = pars, y = vals, group = Chemical)) +
  geom_line()

@@

```{r running-invivoPKfit}
# This is now updated to use dawson chem table
minimal_pk <- pk(data = cvt %>%
                   filter(analyte_dtxsid %in% rat_pars_rat$param_df$Chemical,
                          species %in% "rat"))

@@

new_pk <- do_fit(new_pk, n_cores = 12)
gc()

@@

Here we begin the analysis by estimating the relative likelihood of
"clearance aware" models. This provides an additional evaluation of the evidence
that points towards a restrictive or non-restrictively cleared model.

@@

rel_ll %>%
  rowwise() %>% 
  mutate(across(contains("_1comp"),
                \(x) {
                  ifelse(
                    x < model_flat | x < 1,
                    NA_real_,
                    x
                  )
                })) %>% 
  # drop_na(model_1comp_fup) %>%
  filter(Chemical %in% clear_chem_levels) %>% View()
  glimpse()

@@

new_pk$fit %>%
  filter(param_name %in% "Fup",
         model %in% "model_1comp_fup") %>%
  distinct(Chemical, Species, method, model,
           param_name, estimate, start) %>%
  mutate(direction = ifelse(abs(estimate - start) < 0.1,
                            "similar (less than 10% difference)",
                            ifelse(estimate > start,
                                   "increase",
                                   "decrease"))) %>%
  filter(Chemical %in% clear_chem_levels, direction %in% "similar (less than 10% difference)")
  pivot_longer(cols = c(estimate, start),
               names_to = "parameterization",
               values_to = "value") %>%
  mutate(parameterization = factor(parameterization,
                                   levels = c("start", "estimate"))) %>%
  ggplot(aes(x = parameterization, y = value)) +
  geom_point() +
  geom_line(aes(group = Chemical)) +
  facet_grid(cols = vars(direction)) +
  labs(title = "Optimization of fraction unbound in plasma") +
  theme_bw()

ggsave(filename = paste0(
  Sys.getenv("FIG_DIR"), "RestrictiveClearance_Analysis/",
  today, "fup_optimization_generalFits.png"),
  height = 5, width = 7)

@@

```{r new_analysis}
# Loading the new data

@@

## New reformulated models that estimate Fup!

@@

# Calculating uncertainty of parameter estimation

Below I will use the `coef_sd` method to estimate standard deviation of
the fraction unbound parameter and assess whether this may contribute
to the uncertainty of 'declaration' of clearance for chemicals that have
CvT data documented.

```{r uncertainty_estimation}
this_coefsd <- coef_sd(new_pk)

this_coefsd %>% filter(param_name == "Clint", Chemical %in% clear_chem_levels) %>%
  ggplot(aes(x = factor(Chemical, levels = clear_chem_levels), 
             y = param_value,
             color = model)) +
  geom_point(alpha = 0.5) +
  geom_linerange(aes(ymin = param_value-param_sd,
                    ymax = param_value+param_sd),
                 alpha = 0.5) +
  theme_bw() +
  coord_flip() +
  scale_y_log10()

ggsave(
  filename = paste0(
    Sys.getenv("FIG_DIR"),"RestrictiveClearance_Analysis/",
    today,
    "fup_sd_estimates.png"
  ),
  height = 8,
  width = 8,
  device = "png"
)

this_loglikelihood <- AIC(new_pk) %>%
  filter(Chemical %in% clear_chem_levels) %>%
  mutate(Chemical = factor(Chemical, levels = clear_chem_levels))

this_loglikelihood %>%
  ggplot(aes(x = model, y = Chemical, fill = log10(AIC))) +
  geom_tile() +
  scale_fill_viridis_c(option = "mako", direction = -1) +
  coord_fixed() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

ggsave(
  filename = paste0(
    Sys.getenv("FIG_DIR"),"RestrictiveClearance_Analysis/",
    today,
    "fup_AIC_heatmap.png"
  ),
  height = 10,
  width = 4,
  device = "png"
)

select_chem_info <- get_chem_info_batch(clear_chem_levels)

select_chem_info %>%
  filter(propertyId == "logkow-octanol-water") %>%
  mutate(dtxsid = factor(dtxsid, levels = clear_chem_levels)) %>%
  ggplot(aes(x = value, y = dtxsid, color = source)) +
  geom_point() +
  labs(title = "logKow") +
  geom_vline(xintercept = 5)

# Also check whether the Fup or CLint are at their respective extremes (1, 0)
# If they are, it will be difficult to tell whether a chemical is cleared in a
# particular way

# Slow PBPK model optimization of fup for more chemicals

There may also be an opportunity to evaluate chemicals without _in vivo_ data
by comparing them to those that do. Effectively, we can look at the uncertainty
in the call between restrictive/non-restrictive and further look at estimates
from PBPK models where $f_{up}$ is optimized and it is run through `solve_pbtk`.

```{r slow_pbpk_optimizations}

```{r}
sessionInfo()
```
